---
title: "Classification with $K > 2$ Classes: Multinomial Logistic Regression and Gradient Tree Boosting"
output:
  pdf_document:
    keep_tex: true
geometry: margin=1.5cm
header-includes:
   - \usepackage{booktabs}
   - \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, eval = TRUE)
```

## Restatement of logistic regression

#### Response variable:

$Y_i = \begin{cases} 1 \text{ if observation i is in class 1} \\ 0 \text{ if observation i is in class 0} \end{cases}$

#### Explanatory variables:

$X_i = (X_{i1}, \ldots, X_{ip})$

#### Model:

\begin{align*}
Y_i | X_i = x_i &\sim \text{Bernoulli}(p(x_i)) \\
p(x_i) = P(Y_i = 1 | X_i = x_i) &= \frac{\exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip})}{1 + \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip})} \\
P(Y_i = 0 | X_i = x_i) &= 1 - P(Y_i = 1 | X_i = x_i) \\
&= \frac{1}{1 + \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip})} \\
&= \frac{\exp(0 + 0 x_{i1} + \cdots + 0 x_{ip})}{\exp(0 + 0 x_{i1} + \cdots + 0 x_{ip}) + \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip})}
\end{align*}

#### Note:

$Y_i \sim \text{Bernoulli}(p(x_i))$ is equivalent to $Y_i \sim \text{Binomial}(n = 1, p(x_i))$.

It just means that $Y_i$ is either 0 or 1, and $P(Y_i = 1) = p(x_i)$.

## Multinomial Distribution

We now have $K$ classes; label them as $1, \ldots, K$ (starting at 1 instead of 0 by convention)

#### Notation: Class membership indicator vector

We have defined $Y_i = k$ if observation $i$ is in class $k$.

It will be helpful to have another way of representing this:

\begin{equation*}
{Y^*_i}^{(k)} = \begin{cases} 1 \text{ if $Y_i = k$} \\ 0 \text{ otherwise} \end{cases}
\end{equation*}

For example, if observation $i$ is in class 1, then we can represent this in two ways:

$Y_i = 1$

$({Y^*_i}^{(1)}, {Y^*_i}^{(2)}, \ldots, {Y^*_i}^{(K)}) = (1, 0, \ldots, 0)$

#### Multinomial and Categorical Distributions

The multinomial distribution represents the situation that:

 * We have $n$ independent trials
 * The class membership probabilities are the same for all trials
 * We observe how many of the $n$ trials are in each class

We have to keep track of how many observations were in each class, and what all of the class probabilities are:

$(X_1, \ldots, X_K) \sim \text{Multinomial}\left(n, (p_1, \ldots, p_K)\right)$

We have the following connections between the different distributions:

 * The categorical distribution is the same as the multinomial, but with $n = 1$.

 * The multinomial is similar to the binomial, but for more than 2 classes.

 * The categorical is similar to the Bernoulli, but for more than 2 classes.

## Multinomial logistic regression

#### Model:

\begin{align*}
({Y^*_i}^{(1)}, \ldots, {Y^*_i}^{(K)}) | X_i = x_i &\sim \text{Categorical}\left((p_1(x_i), \ldots, p_K(x_i))\right) \\
p_1(x_i) = P(Y_i = 1 | X_i = x_i) &= \frac{\exp(\beta^{(1)}_0 + \beta^{(1)}_1 x_{i1} + \cdots + \beta^{(1)}_p x_{ip})}{\exp(\beta^{(1)}_0 + \beta^{(1)}_1 x_{i1} + \cdots + \beta^{(1)}_p x_{ip}) + \cdots + \exp(\beta^{(K)}_0 + \beta^{(K)}_1 x_{i1} + \cdots + \beta^{(K)}_p x_{ip})} \\
p_2(x_i) = P(Y_i = 2 | X_i = x_i) &= \frac{\exp(\beta^{(2)}_0 + \beta^{(2)}_1 x_{i1} + \cdots + \beta^{(2)}_p x_{ip})}{\exp(\beta^{(1)}_0 + \beta^{(1)}_1 x_{i1} + \cdots + \beta^{(1)}_p x_{ip}) + \cdots + \exp(\beta^{(K)}_0 + \beta^{(K)}_1 x_{i1} + \cdots + \beta^{(K)}_p x_{ip})} \\
&\vdots \\
p_K(x_i) = P(Y_i = K | X_i = x_i) &= \frac{\exp(\beta^{(K)}_0 + \beta^{(K)}_1 x_{i1} + \cdots + \beta^{(K)}_p x_{ip})}{\exp(\beta^{(1)}_0 + \beta^{(1)}_1 x_{i1} + \cdots + \beta^{(1)}_p x_{ip}) + \cdots + \exp(\beta^{(K)}_0 + \beta^{(K)}_1 x_{i1} + \cdots + \beta^{(K)}_p x_{ip})}
\end{align*}

Basically, this is an involved way of saying that the probability that $Y_i$ is in class $k$ is 

$$\frac{\exp(\beta^{(k)}_0 + \beta^{(k)}_1 x_{i1} + \cdots + \beta^{(k)}_p x_{ip})}{\exp(\beta^{(1)}_0 + \beta^{(1)}_1 x_{i1} + \cdots + \beta^{(1)}_p x_{ip}) + \cdots + \exp(\beta^{(k)}_0 + \beta^{(k)}_1 x_{i1} + \cdots + \beta^{(k)}_p x_{ip})}$$

Note that the first staterment above is equivalent to

$$({Y^*_i}^{(1)}, \ldots, {Y^*_i}^{(K)}) | X_i = x_i \sim \text{Multinomial}\left(n = 1, (p_1(x_i), \ldots, p_K(x_i))\right)$$

#### Lack of identifiability

 * **Definition:** Parameters in a model are *identifiable* if different values of the model parameters yield different distributions for the observable random variables being modelled.

 * The parameters of multinomial logistic regression as stated above are not identifiable.
    * For example, we could add 5 to each of $\beta^{(1)}_0, \ldots, \beta^{(K)}_0$ and obtain the same class probabilities.

 * Different software packages handle this in different ways.  One strategy, used by `nnet::multinom`, is to fix all coefficients for the first class to 0: $\beta^{(1)}_0 = \beta^{(1)}_1 = \beta^{(1)}_p = 0$

 * If a penalty is used on the coefficients (as in lasso or ridge), the lack of identifiability in the model structure is not a problem.

\newpage

#### Softmax transformation of linear functions of explanator variables

 * **Definition:** The *softmax transformation* is defined by the mapping
$$(a_1, a_2, \ldots, a_K) \mapsto \left( \frac{\exp(a_1)}{\sum_k \exp(a_k)}, \frac{\exp(a_2)}{\sum_k \exp(a_k)}, \ldots, \frac{\exp(a_K)}{\sum_k \exp(a_k)} \right)$$
    * The numbers $a_1, \ldots, a_K$ can be any real number
    * $0 < \frac{\exp(a_j)}{\sum_k \exp(a_k)} < 1$ for each $j = 1, \ldots, K$
    * $\sum_j \frac{\exp(a_j)}{\sum_k \exp(a_k)} = 1$
    * So basically, we took some real numbers $a_1, \ldots, a_K$ (could be negative! could be large!) and turned them into a valid set of probabilities for the $K$ categories.

 * One way of viewing multinomial logistic regression is that we pass $K$ linear regressions through the softmax transformation to obtain a set of class membership probabilities that sum to 1.

## Gradient Tree Boosting for Multiple Classes

#### Main Idea: softmax transformation of sums of regression trees

 * Last class, our big idea was to take logistic regression and replace the (essentially) linear form in the explanatory variables with a sum of regression trees
 * Now we will do the same, but we associate a separate sum of regression trees with each class:

\begin{align*}
P(Y_i = 1 | X_i) &= p_1(x_i) = \frac{\exp(\text{sum of regression trees for class 1})}{\exp(\text{sum of regression trees for class 1}) + \cdots + \exp(\text{sum of regression trees for class K})} \\
 &= \frac{\exp\left\{a^{(1)}(x_i)\right\}}{\exp\left\{a^{(1)}(x_i)\right\} + \cdots + \exp\left\{a^{(K)}(x_i)\right\}} \\
P(Y_i = 2 | X_i) &= p_2(x_i) =  \frac{\exp(\text{sum of regression trees for class 2})}{\exp(\text{sum of regression trees for class 1}) + \cdots + \exp(\text{sum of regression trees for class K})} \\
 &= \frac{\exp\left\{a^{(2)}(x_i)\right\}}{\exp\left\{a^{(1)}(x_i)\right\} + \cdots + \exp\left\{a^{(K)}(x_i)\right\}} \\
&\vdots \\
P(Y_i = K | X_i) &= p_K(x_i) =  \frac{\exp(\text{sum of regression trees for class K})}{\exp(\text{sum of regression trees for class 1}) + \cdots + \exp(\text{sum of regression trees for class K})} \\
 &= \frac{\exp\left\{a^{(K)}(x_i)\right\}}{\exp\left\{a^{(1)}(x_i)\right\} + \cdots + \exp\left\{a^{(K)}(x_i)\right\}}
\end{align*}

where $a^{(k)}(x_i)$ is obtained as a sum of $B$ regression trees associated with class $k$:

$$a^{(k)}(x_i) = \sum_{b = 1}^B \sum_{m = 1}^{\vert T \vert^{(b, k)}} \mathbb{I}_{R_m^{(b, k)}}(x_i) \hat{y}_m^{(b, k)}.$$

\newpage

#### Algorithm: Gradient Tree Boosting for Multiple Classes

The gradient tree boosting procedure estimates the regression trees one group at a time, fitting each tree to (two views of the same thing):

 * The gradient of the log-likelihood, which is based on the model $Y_i \vert X_i \sim \text{Multinomial}\left(n = 1, p = (p_1(x_i), \ldots, p_K(x_i))\right)$
 * The "residuals" ${y^*_i}^{(k)} - p_k(x_i)$

Notes:

 * We could derive these residuals as the derivatives of the log-likelihood
 * If we had 2 classes, this is exactly the same as the procedure we talked about for boosting with 2 classes last week
 * This is like doing a first-order Taylor series approximation to the log-likelihood; xgboost implements a second-order Taylor series approximation.

Algorithm:

 1. Initialize $\hat{a}^{(k, 0)}(x) = 0$ for all $k$.  (first index in superscript is class, second is boosting iteration)
 2. For $b = 1, \ldots, B$:
    a. Compute the predicted probability of being in class k for each observation $i$, based on the ensemble that was created after the previous step: $p^{(b-1)}_k(x_i) = \hat{P}^{(b-1)}(Y_i = k \vert x_i) = \frac{e^{\hat{a}^{(k, b-1)}(x_i)}}{e^{\hat{a}^{(1, b-1)}(x_i)} + \cdots + e^{\hat{a}^{(K, b-1)}(x_i)}}$
    b. For each class $k = 1, \ldots, K$, fit a regression tree $\hat{g}^{(k, b)}(x)$ using the residuals ${y^*_i}^{(k)} - p^{(b-1)}_k(x_i)$ as the response.
    c. Update the ensemble by adding in this new model: $\hat{a}^{(k, b)}(x) = \hat{a}^{(k, b-1)}(x) + \hat{g}^{(k, b)}(x)$

## Example: Vertebral Column

Our data example uses "six biomechanical attributes derived from the shape and orientation of the pelvis and lumbar spine" for a patient to classify the patient into one of three groups representing different conditions that may be affecting their spine: "DH" for disk hernia, "SL" for Spondylolisthesis, or "NO" for normal (neither of the other two conditions).

The data are available at https://archive.ics.uci.edu/ml/datasets/Vertebral+Column and were discussed in the following article:

Berthonnaud, E., Dimnet, J., Roussouly, P. & Labelle, H. (2005). 'Analysis of the sagittal balance of the spine and pelvis using shape and orientation parameters', Journal of Spinal Disorders & Techniques, 18(1):40–47. 

#### Reading the data in, preprocessing, train/test split, cross-validation splits

```{r, message = FALSE}
library(readr)
library(purrr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(rpart)
library(caret)

vertebral_column <- read_table2("http://www.evanlray.com/data/UCIML/vertebral_column/column_3C.dat",
  col_names = FALSE)
names(vertebral_column) <- c(paste0("X_", 1:6), "type")

vertebral_column <- vertebral_column %>%
  mutate(
    type = factor(type)
  )

set.seed(723)

# Train/test split
tt_inds <- caret::createDataPartition(vertebral_column$type, p = 0.7)
train_set <- vertebral_column %>% slice(tt_inds[[1]])
test_set <- vertebral_column %>% slice(-tt_inds[[1]])

# Cross-validation splits
crossval_val_fold_inds <- caret::createFolds(
  y = train_set$type, # response variable as a vector
  k = 10 # number of folds for cross-validation
)

get_complementary_inds <- function(val_inds) {
  return(seq_len(nrow(train_set))[-val_inds])
}

crossval_train_fold_inds <- map(crossval_val_fold_inds, get_complementary_inds)

# Cross-validation splits
crossval_val_fold_inds <- caret::createFolds(
  y = train_set$type, # response variable as a vector
  k = 10 # number of folds for cross-validation
)

get_complementary_inds <- function(val_inds) {
  return(seq_len(nrow(train_set))[-val_inds])
}

crossval_train_fold_inds <- map(crossval_val_fold_inds, get_complementary_inds)
```


#### Fit and test set classification error rate via multinomial logistic regression

Fit model, pick tuning parameter values yielding highest cross-validated accuracy.

```{r}
multilogistic_fit <- train(
  type ~ .,
  data = train_set,
  trace = FALSE,
  method = "multinom",
  trControl = trainControl(
    method = "cv",
    number = 10,
    returnResamp = "all",
    index = crossval_train_fold_inds,
    indexOut = crossval_val_fold_inds,
    savePredictions = TRUE 
  ),
  tuneGrid = data.frame(decay = seq(from = 0, to = 0.2, length = 30))
)

multilogistic_fit$results %>% filter(Accuracy == max(Accuracy))
```

Test set performance: looking for low test set error rate, high test set accuracy

```{r}
mean(test_set$type != predict(multilogistic_fit, test_set))
mean(test_set$type == predict(multilogistic_fit, test_set))
```

\newpage

#### Fit and test set classification error rate via gradient tree boosting

Fit model, pick tuning parameter values yielding highest cross-validated accuracy.

```{r}
xgb_fit <- train(
  type ~ .,
  data = train_set,
  method = "xgbTree",
  trControl = trainControl(
    method = "cv",
    number = 10,
    returnResamp = "all",
    index = crossval_train_fold_inds,
    indexOut = crossval_val_fold_inds,
    savePredictions = TRUE 
  ),
  tuneGrid = expand.grid(
    nrounds = c(5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
    eta = c(0.5, 0.6, 0.7), # learning rate; 0.3 is the default
    gamma = 0, # minimum loss reduction to make a split; 0 is the default
    max_depth = 1:5, # how deep are our trees?
    subsample = c(0.5, 0.9, 1), # proportion of observations to use in growing each tree
    colsample_bytree = 1, # proportion of explanatory variables used in each tree
    min_child_weight = 1 # think of this as how many observations must be in each leaf node
  )
)

xgb_fit$results %>% filter(Accuracy == max(Accuracy))
```

Test set performance: looking for low test set error rate, high test set accuracy

```{r}
mean(test_set$type != predict(xgb_fit, test_set))
mean(test_set$type == predict(xgb_fit, test_set))
```


#### Fit and test set classification error rate via K nearest neighbors

Fit model, pick tuning parameter values yielding highest cross-validated accuracy.

```{r}
knn_fit <- train(
  type ~ .,
  data = train_set,
  method = "knn",
  trControl = trainControl(
    method = "cv",
    number = 10,
    returnResamp = "all",
    index = crossval_train_fold_inds,
    indexOut = crossval_val_fold_inds,
    savePredictions = TRUE 
  ),
  tuneLength = 5
)

knn_fit$results %>% filter(Accuracy == max(Accuracy))
```

Test set performance: looking for low test set error rate, high test set accuracy

```{r}
mean(test_set$type != predict(knn_fit, test_set))
mean(test_set$type == predict(knn_fit, test_set))
```


#### Fit and test set classification error rate via random forest

Fit model, pick tuning parameter values yielding highest cross-validated accuracy.

```{r}
rf_fit <- train(
  type ~ .,
  data = train_set,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10,
    returnResamp = "all",
    index = crossval_train_fold_inds,
    indexOut = crossval_val_fold_inds,
    savePredictions = TRUE 
  ),
  tuneLength = 5
)

rf_fit$results %>% filter(Accuracy == max(Accuracy))
```

Test set performance: looking for low test set error rate, high test set accuracy

```{r}
mean(test_set$type != predict(rf_fit, test_set))
mean(test_set$type == predict(rf_fit, test_set))
```











