---
title: "Multiple Regression with both Categorical and Quantitative Explanatory Variables"
output: pdf_document
geometry: margin=1.5cm
classoption: landscape
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Recall that we are thinking about a data set with several variables recorded about 1753 movies.  We are exploring building multiple regression models for a movie's international gross earnings in inflation-adjusted 2013 dollars (`intgross_2013`) based on the following 5 explanatory variables:

 1. `budget_2013` (quantitative)
 2. `run_time_min` (quantitative)
 3. `imdb_rating` (quantitative)
 4. `mpaa_rating` (categorical)
 5. `bechdel_test_binary` (categorical)

First, let's load the data in, filter to include only MPAA ratings categories with a reasonable number of movies in them, set categorical variables to factors, and apply a log transformation to `intgross_2013`, `budget_2013`, and `run_time_min`.

```{r, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2) # general plotting functionality
library(GGally) # includes the ggpairs function, pairs plots via ggplot2
library(gridExtra) # for grid.arrange, which arranges the plots next to each other

options(na.action = na.exclude, digits = 7)

movies <- read_csv("http://www.evanlray.com/data/bechdel/bechdel.csv") %>%
  filter(mpaa_rating %in% c("G", "PG", "PG-13", "R")) %>%
  mutate(
    bechdel_test = factor(bechdel_test, levels = c("nowomen", "notalk", "men", "dubious", "ok")),
    bechdel_test_binary = factor(bechdel_test_binary, levels = c("FAIL", "PASS")),
    mpaa_rating = factor(mpaa_rating, levels = c("G", "PG", "PG-13", "R"))
  ) %>%
  mutate(
    log_intgross_2013 = log(intgross_2013),
    log_budget_2013 = log(budget_2013),
    log_run_time_min = log(run_time_min)
  )
```

Our goals:

 * Understand R's parameterization of linear models involving categorical variables (interpretation of fixed and interaction effects)
 * See some examples of testing effects

\newpage

# 1 categorical explanatory variable (one-way ANOVA)

Here are plots showing the relationship between the categorical variables and the response:

```{r}
p_mpaa <- ggplot(data = movies, mapping = aes(x = log_intgross_2013, color = mpaa_rating, linetype = mpaa_rating)) +
  geom_density()

p_bechdel <- ggplot(data = movies,
  mapping = aes(x = log_intgross_2013, color = bechdel_test_binary, linetype = bechdel_test_binary)) +
  geom_density()

grid.arrange(p_mpaa, p_bechdel, ncol = 1)
```

## Explanatory variable has 2 categories

```{r}
movies <- movies %>% filter(!is.na(bechdel_test_binary))
fit_bechdel <- lm(log_intgross_2013 ~ bechdel_test_binary, data = movies)
summary(fit_bechdel)
```

#### What is the estimated equation for this fit?  Define all variables involved.

\newpage

#### How do the parameter estimates relate to the following R output?

```{r}
group_means <- movies %>%
  group_by(bechdel_test_binary) %>%
  summarize(
    mean_log_earnings = mean(log_intgross_2013, na.rm = TRUE)
  ) %>%
  as.data.frame()
group_means

levels(movies$bechdel_test_binary)
```

\vspace{1cm}

## Explanatory variable has >2 categories

```{r}
fit_mpaa <- lm(log_intgross_2013 ~ mpaa_rating, data = movies)
summary(fit_mpaa)
```


#### What is the estimated equation for this fit?  Define all variables involved.

\vspace{2cm}

#### How do the parameter estimates relate to the following R output?

```{r}
group_means <- movies %>%
  group_by(mpaa_rating) %>%
  summarize(
    mean_log_earnings = mean(log_intgross_2013, na.rm = TRUE)
  ) %>%
  as.data.frame()
group_means

levels(movies$mpaa_rating)
```

\newpage

## A Cautionary Tale - Regression with Ordered Factors

```{r}
movies <- movies %>%
  mutate(
    bechdel_test_binary_ordered_factor = factor(bechdel_test_binary, levels = c("FAIL", "PASS"), ordered = TRUE)
  )

fit_bechdel_ordered_factor <- lm(log_intgross_2013 ~ bechdel_test_binary_ordered_factor, data = movies)
summary(fit_bechdel_ordered_factor)

movies %>%
  group_by(bechdel_test_binary) %>%
  summarize(
    mean_log_earnings = mean(log_intgross_2013, na.rm = TRUE)
  ) %>%
  as.data.frame()
```

\newpage

# Both Categorical and Quantitative Variables; no interactions

First, we subset to movies where all of our candidate explanatory variables are non-missing.  This is necessary to ensure that all fits below are based on the same observations, which is needed for comparing models with `anova`.

```{r}
movies <- movies %>%
  filter(!is.na(log_intgross_2013) & !is.na(mpaa_rating) & !is.na(bechdel_test_binary) &
           !is.na(log_budget_2013) & !is.na(log_run_time_min) & !is.na(imdb_rating))
```

Let's try a backwards selection type strategy: we'll use all the explanatory variables we're considering, then drop variables that don't seem to be contributing much to the fit.

```{r}
fit_all_x <- lm(log_intgross_2013 ~ mpaa_rating + bechdel_test_binary + log_budget_2013 + log_run_time_min + imdb_rating,
  data = movies)
summary(fit_all_x)
```

This initial fit suggests that after accounting for the associations between all other explanatory variables and log earnings, a movie's run time and whether or not it passes the Bechdel test do not account for a statistically significant amount of variation in earnings.

Let's conduct an F test to see whether we might drop both of these variables from the model.  We fit a reduced model and compare with `anova`:

```{r}
fit_mpaa_budget_imdb <- lm(log_intgross_2013 ~ mpaa_rating + log_budget_2013 + imdb_rating,
  data = movies)
anova(fit_all_x, fit_mpaa_budget_imdb)
```

#### What is the result of the test?

\vspace{2cm}

#### In the reduced model fit above, what's the interpretation of the estimated coefficient for `mpaa_ratingR`?

\vspace{2cm}

\newpage

# Interactions between quantitative and categorical variables

Here's a model fit that includes an interaction between `log_budget_2013` and `mpaa_rating`, as well as a call to `anova` that compares this model with the previous model that did not include interactions.

```{r}
fit_interaction <- lm(log_intgross_2013 ~ mpaa_rating + log_budget_2013 + imdb_rating + mpaa_rating:log_budget_2013,
  data = movies)
summary(fit_interaction)
anova(fit_interaction, fit_mpaa_budget_imdb)
```

#### What's the estimated equation for this model?

\vspace{2cm}

#### In the model fit including interactions, what is the interpretation of the estimated coefficient for `mpaa_ratingPG-13:log_budget_2013`?

\vspace{2cm}

#### According to the hypothesis test, should we include the interaction in the model?


