---
title: "Polynomial Regression"
author: "Adapted from De Veaux, Velleman, and Bock"
date: "September 07, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Emperor penguins can slow their heart rates while diving.  Here's a plot showing 125 observations of penguin dives, with the duration of the penguin's dive on the horizontal axis and the penguin's heart rate on the vertical axis.

```{r, fig.height = 3, fig.width = 3, message=FALSE, warning=FALSE}
library(readr) # for read_csv, which can read csv files from the internet
library(dplyr) # for data manipulation functions
library(ggplot2) # for making plots

penguins <- read_csv("http://www.evanlray.com/data/sdm4/Penguins.csv")

ggplot() + 
  geom_point(data = penguins, mapping = aes(x = Duration, y = DiveHeartRate))
```

## Linear Fit

Is a simple linear regression model good enough?  Let's fit a model and look at some diagnostic plots to find out:

```{r, fig.height = 3.75}
slr_fit <- lm(DiveHeartRate ~ Duration, data = penguins)
predict_slr <- function(x) {
  predict(slr_fit, data.frame(Duration = x))
}

ggplot(data = penguins, mapping = aes(x = Duration, y = DiveHeartRate)) + 
  geom_point() +
  stat_function(fun = predict_slr) +
  ggtitle("Linear fit")

penguins <- penguins %>%
  mutate(
    residual_slr = residuals(slr_fit)
  )

ggplot(data = penguins, mapping = aes(x = Duration, y = residual_slr)) +
  geom_point() + 
  geom_smooth() +
  ggtitle("Residuals vs Duration diagnostic plot, linear fit")
```

There is a clear trend in the residuals.  Let's try fitting a parabola instead.

## Quadratic Fit

Note the addition of ` + I(Duration^2)` in the model formula.

```{r, fig.height = 3.75}
quad_fit <- lm(DiveHeartRate ~ Duration + I(Duration^2), data = penguins)
summary(quad_fit)

predict_quad <- function(x) {
  predict(quad_fit, data.frame(Duration = x))
}

ggplot(data = penguins, mapping = aes(x = Duration, y = DiveHeartRate)) + 
  geom_point() +
  stat_function(fun = predict_quad) +
  ggtitle("Quadratic fit")

penguins <- penguins %>%
  mutate(
    residual_quad = residuals(quad_fit)
  )

ggplot(data = penguins, mapping = aes(x = Duration, y = residual_quad)) +
  geom_point() + 
  geom_smooth() +
  ggtitle("Residuals vs Duration diagnostic plot, quadratic fit")
```

Not as much of a trend...  What happens if we fit a cubic polynomial?

\newpage

## Cubic Fit

```{r, fig.height = 3.75}
cubic_fit <- lm(DiveHeartRate ~ Duration + I(Duration^2) + I(Duration^3), data = penguins)
summary(cubic_fit)

predict_cubic <- function(x) {
  predict(cubic_fit, data.frame(Duration = x))
}

ggplot(data = penguins, mapping = aes(x = Duration, y = DiveHeartRate)) + 
  geom_point() +
  stat_function(fun = predict_cubic) +
  ggtitle("cubic fit")

penguins <- penguins %>%
  mutate(
    residual_cubic = residuals(cubic_fit)
  )

ggplot(data = penguins, mapping = aes(x = Duration, y = residual_cubic)) +
  geom_point() + 
  geom_smooth() +
  ggtitle("Residuals vs Duration diagnostic plot, cubic fit")
```

We can also get the same model fit another way, using `poly()` instead of `I()`:

```{r}
cubic_fit <- lm(DiveHeartRate ~ poly(Duration, 3), data = penguins)
summary(cubic_fit)

```

\newpage

## Another Approach... Data Transformation

```{r, fig.height = 3.75}
log_fit <- lm(log(DiveHeartRate) ~ Duration, data = penguins)
predict_log <- function(x) {
  predict(log_fit, data.frame(Duration = x))
}

ggplot(data = penguins, mapping = aes(x = Duration, y = log(DiveHeartRate))) + 
  geom_point() +
  stat_function(fun = predict_log) +
  ggtitle("Linear fit to log(DiveHeartRate)")

penguins <- penguins %>%
  mutate(
    residual_log = residuals(log_fit)
  )

ggplot(data = penguins, mapping = aes(x = Duration, y = residual_log)) +
  geom_point() + 
  geom_smooth() +
  ggtitle("Residuals vs Duration diagnostic plot, log fit")
```
