---
title: "Multiple Regression with Quantitative Explanatory Variables"
date: "September 12, 2018"
output: pdf_document
geometry: margin=1.5cm
classoption: landscape
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Recall that we are thinking about a data set with several variables recorded about 1753 movies.  Over the next week or so, we will explore building multiple regression models for a movie's international gross earnings in inflation-adjusted 2013 dollars (`intgross_2013`) based on the following 5 explanatory variables:

 1. `budget_2013`
 2. `run_time_min`
 3. `imdb_rating`
 4. `mpaa_rating`
 5. `bechdel_test_binary`

Today, we'll look at just a few models based on the three quantitative explanatory variables, `budget_2013`, `run_time_min`, and `imdb_rating`.

First, let's load the data in, filter to include only MPAA ratings categories with a reasonable number of movies in them, and set categorical variables to factors.  If anything in the following R code isn't clear to you, you should ask about it.

```{r, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2) # general plotting functionality
library(GGally) # includes the ggpairs function, pairs plots via ggplot2
library(gridExtra) # for grid.arrange, which arranges the plots next to each other

options(na.action = na.exclude)


movies <- read_csv("http://www.evanlray.com/data/bechdel/bechdel.csv") %>%
  filter(mpaa_rating %in% c("G", "PG", "PG-13", "R")) %>%
  mutate(
    bechdel_test = factor(bechdel_test, levels = c("nowomen", "notalk", "men", "dubious", "ok"), ordered = TRUE),
    bechdel_test_binary = factor(bechdel_test_binary, levels = c("FAIL", "PASS"), ordered = TRUE),
    mpaa_rating = factor(mpaa_rating, levels = c("G", "PG", "PG-13", "R"), ordered = TRUE)
  )
```

Last class we looked at a pairs plot and decided some transformations were needed to make a linear model feasible.  Neither a log transformation nor a square root transformation were great, but a log transformation will be good enough for our explorations today (again, we'll return to a more complete discussion of transformations later).

```{r, warning=FALSE, message=FALSE, fig.height=4.5, fig.width=10}
movies <- movies %>%
  mutate(
    log_intgross_2013 = log(intgross_2013),
    log_budget_2013 = log(budget_2013),
    log_run_time_min = log(run_time_min)
  )

vars_to_use <- c("log_budget_2013", "log_run_time_min", "imdb_rating", "log_intgross_2013")
ggpairs(movies[, vars_to_use])
```

For today, let's fit a few quick linear models based on the log transformation.

There are 3 main things to get out of this:

1. Coefficient estimates, interpretations, and hypothesis test results for a variable depend on what other variables are in the model
2. Multiple linear regression can be used to fit surfaces other than planes (i.e., curved surfaces).  We'll see how to do this with either:
    a. higher-degree terms in one of the explanatory variables
    b. interactions between two explanatory variables
3. t tests (about one coefficient's value) vs. F tests (simultaneous test about the values of multiple coefficients)

\newpage

# Concept 1: Coefficient estimates, interpretations, and hypothesis test results for a variable depend on what other variables are in the model!

### Fit 1: Explanatory variable is run time

```{r}
fit_run_time <- lm(log_intgross_2013 ~ log_run_time_min, data = movies)
summary(fit_run_time)
```

* Does a movie's (log) run time explain a statistically significant amount of variation in a movie's earnings?  Conduct a hypothesis test.

\vspace{3cm}

* Interpret the coefficient estimate for (log) run time.

\newpage

### Fit 2: Explanatory variables are budget, imdb rating, and run time

```{r}
fit_all_x <- lm(log_intgross_2013 ~ log_budget_2013 + imdb_rating + log_run_time_min, data = movies)
summary(fit_all_x)
```

* Does a movie's (log) run time explain a statistically significant amount of variation in a movie's earnings after accounting for the linear associations between log budget, imdb rating, and log earnings?  Conduct a hypothesis test.

\vspace{1.75cm}

* Interpret the coefficient estimate for (log) run time

\vspace{1.75cm}

* Compare what we learned about the value of run time for predicting earnings from fits 1 and 2.  Are the findings consistent?

\newpage

# Concept 2: Surfaces other than planes!

First, let's verify that a "standard" multiple linear regression fit gives us a plane.  Then we'll see two specific examples of fitting something other than a plane.

### Fit 3: Explanatory variables are budget and IMDB rating - fitting a plane

```{r}
fit_budget_imdb <- lm(log_intgross_2013 ~ log_budget_2013 + imdb_rating, data = movies)
summary(fit_budget_imdb)
```

 * What's the equation of the estimated surface?
 
\newpage
 
Here's a plot of our estimated surface and the data points (I will never ask you to make a plot like this):

```{r}
log_budget_grid <- seq(from = min(movies$log_budget_2013), to = max(movies$log_budget_2013), length = 21)
imdb_grid <- seq(from = min(movies$imdb_rating), to = max(movies$imdb_rating), length = 21)

f = outer(log_budget_grid, imdb_grid, function(log_budget, imdb) {
  predict(fit_budget_imdb, data.frame(log_budget_2013 = log_budget, imdb_rating = imdb))
})

nrz <- nrow(f)
ncz <- ncol(f)
# Create a function interpolating colors in the range of specified colors
jet.colors <- colorRampPalette( c("blue", "red") )
# Generate the desired number of colors from this palette
nbcol <- 100
color <- jet.colors(nbcol)
# Compute the z-value at the facet centres
ffacet <- f[-1, -1] + f[-1, -ncz] + f[-nrz, -1] + f[-nrz, -ncz]
# Recode facet z-values into color indices
facetcol <- cut(ffacet, nbcol)

res <- persp(log_budget_grid, imdb_grid, f, col = color[facetcol], theta = 40, phi = 40)
points(trans3d(movies$log_budget_2013, movies$imdb_rating, movies$log_intgross_2013, pmat = res), col = 1, pch = 16)
```

OK... let's look at residual diagnostic plots for this model.

\newpage

```{r, fig.width=10, fig.height = 3, warning=FALSE, message=FALSE}
movies <- movies %>%
  mutate(
    residuals_budget_imdb = residuals(fit_budget_imdb),
    predicted_budget_imdb = predict(fit_budget_imdb)
  )

p1 <- ggplot(data = movies, mapping = aes(x = log_budget_2013, y = residuals_budget_imdb)) +
  geom_point() +
  geom_smooth()

p2 <- ggplot(data = movies, mapping = aes(x = imdb_rating, y = residuals_budget_imdb)) +
  geom_point() +
  geom_smooth()

p3 <- ggplot(data = movies, mapping = aes(x = predicted_budget_imdb, y = residuals_budget_imdb)) +
  geom_point() +
  geom_smooth()

grid.arrange(p1, p2, p3, nrow = 1)
```

The residual diagnostic plots give an appearance of a lack of linearity in the effect of `log_budget_2013` on `log_intgross_2013`.  This is driven by a small number of movies with low budgets.  It's not clear to me whether our sample size in that region is really large enough to give strong/reliable evidence about a non-linear effect.  However, let's consider a fit with a quadratic term in `log_budget_2013`.

## Concept 2 (a): Adding polynomial terms in one variable

### Fit 4: Explanatory variables are budget (quadratic effect) and IMDB rating (linear effect)

```{r, fig.width=10, fig.height = 3, warning=FALSE, message = FALSE}
fit_budget_sq_imdb <- lm(log_intgross_2013 ~ log_budget_2013 + I(log_budget_2013^2) + imdb_rating, data = movies)
summary(fit_budget_sq_imdb)

movies <- movies %>%
  mutate(
    residuals_budget_sq_imdb = residuals(fit_budget_sq_imdb),
    predicted_budget_sq_imdb = predict(fit_budget_sq_imdb)
  )

p1 <- ggplot(data = movies, mapping = aes(x = log_budget_2013, y = residuals_budget_sq_imdb)) +
  geom_point() +
  geom_smooth()

p2 <- ggplot(data = movies, mapping = aes(x = imdb_rating, y = residuals_budget_sq_imdb)) +
  geom_point() +
  geom_smooth()

p3 <- ggplot(data = movies, mapping = aes(x = predicted_budget_imdb, y = residuals_budget_sq_imdb)) +
  geom_point() +
  geom_smooth()

grid.arrange(p1, p2, p3, nrow = 1)
```

 * What is the equation of the estimated surface?
 
\vspace{2cm} 
 

The coefficient for `I(log_budget_2013^2)` describes the curvature of the fitted surface along the `log_budget_2013` axis.

Here is a view of the resulting fitted surface.

```{r}
log_budget_grid <- seq(from = min(movies$log_budget_2013), to = max(movies$log_budget_2013), length = 21)
imdb_grid <- seq(from = min(movies$imdb_rating), to = max(movies$imdb_rating), length = 21)

f = outer(log_budget_grid, imdb_grid, function(log_budget, imdb) {
  predict(fit_budget_sq_imdb, data.frame(log_budget_2013 = log_budget, imdb_rating = imdb))
})

nrz <- nrow(f)
ncz <- ncol(f)
# Create a function interpolating colors in the range of specified colors
jet.colors <- colorRampPalette( c("blue", "red") )
# Generate the desired number of colors from this palette
nbcol <- 100
color <- jet.colors(nbcol)
# Compute the z-value at the facet centres
ffacet <- f[-1, -1] + f[-1, -ncz] + f[-nrz, -1] + f[-nrz, -ncz]
# Recode facet z-values into color indices
facetcol <- cut(ffacet, nbcol)

res <- persp(log_budget_grid, imdb_grid, f, col = color[facetcol], theta = 10, phi = 10)
points(trans3d(movies$log_budget_2013, movies$imdb_rating, movies$log_intgross_2013, pmat = res), pch = 16)
```

## Concept 2 (b): Adding interactions between two explanatory variables

### Fit 5: Interaction between budget and run time

```{r, fig.width=10}
fit_budget_runtime_interaction <- lm(log_intgross_2013 ~ log_budget_2013 * run_time_min, data = movies)
summary(fit_budget_runtime_interaction)
```

 * What is the equation of the estimated surface?
 
 \vspace{2cm}
 
 * What is the interpretation of the coefficient for the interaction?
 
 \vspace{2cm}
 
```{r}
log_budget_grid <- seq(from = min(movies$log_budget_2013), to = max(movies$log_budget_2013), length = 21)
run_time_grid <- seq(from = min(movies$run_time_min, na.rm = TRUE), to = max(movies$run_time_min, na.rm = TRUE), length = 21)

f = outer(log_budget_grid, run_time_grid, function(log_budget, run_time) {
  predict(fit_budget_runtime_interaction, data.frame(log_budget_2013 = log_budget, run_time_min = run_time))
})

nrz <- nrow(f)
ncz <- ncol(f)
# Create a function interpolating colors in the range of specified colors
jet.colors <- colorRampPalette( c("blue", "red") )
# Generate the desired number of colors from this palette
nbcol <- 100
color <- jet.colors(nbcol)
# Compute the z-value at the facet centres
ffacet <- f[-1, -1] + f[-1, -ncz] + f[-nrz, -1] + f[-nrz, -ncz]
# Recode facet z-values into color indices
facetcol <- cut(ffacet, nbcol)

res <- persp(log_budget_grid, run_time_grid, f, col = color[facetcol], theta = 20, phi = 0)

res <- persp(log_budget_grid, run_time_grid, f, col = color[facetcol], theta = 10, phi = 15)
```

# Concept 3: t tests (one coefficient) vs. F tests (multiple coefficients simultaneously); groups of individually non-significant terms can be jointly significant

### Fit 6: Interaction between budget and imdb rating

```{r, fig.width=10}
fit_budget_imdb_interaction <- lm(log_intgross_2013 ~ log_budget_2013 * imdb_rating, data = movies)
summary(fit_budget_imdb_interaction)

fit_budget_only <- lm(log_intgross_2013 ~ log_budget_2013, data = movies)
anova(fit_budget_imdb_interaction, fit_budget_only)
```

 * Conduct a hypothesis test of the claim that after accounting for the linear effect of `log_budget_2013` and the interaction effect between `log_budget_2013` and `imdb_rating`, there is no linear association between `imdb_rating` and `log_intgross_2013`.
 
 \vspace{3cm}
 
 * Conduct a hypothesis test of the claim that after accounting for the linear effect of `log_budget_2013` and the linear effect of `imdb_rating`, there is no interaction effect of `log_budget_2013` and `imdb_rating`, there is no linear association between `imdb_rating` and `log_intgross_2013`.
 
 \vspace{3cm}
 
 * Conduct a hypothesis test of the claim that after accounting for the linear effect of `log_budget_2013`, there is no linear effect of `imdb_rating` and no interaction effect between `imdb_rating` and `log_budget_2013`.
 
 \vspace{3cm}
 
* Conduct a hypothesis test of the claim that there is no linear effect of either `log_budget_2013` or `imb_rating`, and there is also no interaction effect between `imdb_rating` and `log_budget_2013`.
